---
title: "Assignment 3"
author: "Rory Sarten 301005654"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
```

## Question 1

a)

A Runs Test tests a set of binary variables $X_1, ..., X_n$ to verify if the variables occur randomly.

$H_0$: variables occur randomly, i.e. knowing $X_1, ..., X_n$ does not help predict $X_{n+1}$.

$H_A$: variables are not random, i.e. knowing some part of the sequence can help predict subsequent variables.

As the variables are binary, they will take the value $0$ or $1$. The number of $0$s is $n_0$ and the number of $1$s is $n_1$, where:

$n_0 = n -\sum\limits_{i=1}^nX_i$

$n_1 = \sum\limits_{i=1}^nX_i$

To perform a Runs Test the observations are combined into one collection of $n = n_0 + n_1$ observations and arranged in increasing order of magnitude or observation. They are labeled according to which set they originally came from. A run is a group of two or more sequential values of $0$ or $1$.

Let $R$ denote the number of runs in the combined ordered sample of $X \in \{0, 1\}$. Under $H_0$, $R$ can be approximated as a normally distributed random variable, assuming both $n_0$ and $n_1$ are sufficiently large.

$\bar{R} = \dfrac{2n_0n_1}{n} + 1$

$Var(\bar{R}) = \dfrac{2n_0n_1(2n_0n_1-n)}{n^2(n - 1)}$

With test statistic $Z = \dfrac{R-\bar{R}}{\sqrt{Var(\bar{R})}}$ where $Z \sim N(0, 1)$

b)

```{r runs_test}
## 0 healthy, 1 has disease
X <- "HHHDDDHDDHHHHHDHHHHHHHHHHHDDDHHHHDDDHHHHDHHHDHH" %>% 
  stringr::str_split("") %>% unlist() %>% `==`("D") %>% as.integer()

n_0 <- length(X) - sum(X)
n_1 <- sum(X)

(2 * n_0 * n_1)/length(X) + 1


```