---
title: "Assignment 3"
author: "Rory Sarten 301005654"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(ggplot2)
library(gghighlight)
```

## Question 1

a)

A Runs Test tests a set of binary variables $X_1, ..., X_n$ to verify if the variables occur randomly.

$H_0$: variables occur randomly, i.e. knowing $X_1, ..., X_n$ does not help predict $X_{n+1}$.

$H_A$: variables are not random, i.e. knowing some part of the sequence can help predict subsequent variables.

As the variables are binary, they will take the value $0$ or $1$. The number of $0$s is $n_0$ and the number of $1$s is $n_1$, where:

$n_0 = n -\sum\limits_{i=1}^nX_i$

$n_1 = \sum\limits_{i=1}^nX_i$

To perform a Runs Test the observations are combined into one collection of $n = n_0 + n_1$ observations and arranged in increasing order of magnitude or observation. They are labeled according to which set they originally came from. A run is a group of two or more sequential values of $0$ or $1$.

Let $R$ denote the number of runs in the combined ordered sample of $X \in \{0, 1\}$. Under $H_0$, $R$ can be approximated as a normally distributed random variable, assuming both $n_0$ and $n_1$ are sufficiently large.

$R = 1 + \sum\limits_{i=2}^nI_{(X_i, X_{i-1})}$, where $I_{(X_i, X_{i-1})} = 0$ if $X_i = X_{i-1}$ and $I_{(X_i, X_{i-1})} = 1$ if $X_i \ne X_{i-1}$

$\bar{R} = \dfrac{2n_0n_1}{n} + 1$

$Var(\bar{R}) = \dfrac{2n_0n_1(2n_0n_1-n)}{n^2(n - 1)}$

With test statistic $Z = \dfrac{R-\bar{R}}{\sqrt{Var(\bar{R})}}$ where $Z \sim N(0, 1)$

b)

A small number of runs (a small value for $R$) would indicate that $X_i$ is more likely to be the same as $X_{i-1}$. A large number means that $X$ is fluctuating regularly between values and $X$ is less likely to be the same as $X_{i - 1}$.

c)

```{r runs_test_parametric}
## 0 healthy, 1 has disease
X <- "HHHDDDHDDHHHHHDHHHHHHHHHHHDDDHHHHDDDHHHHDHHHDHH" %>% 
  stringr::str_split("") %>% unlist() %>% `==`("D") %>% as.integer()

n <- length(X)
n_0 <- n - sum(X)
n_1 <- sum(X)

R <- 1 + sum(X[2:n] != X[1:n-1])
R_est <- 1 + 2*n_0*n_1/n
R_var <- (2*n_0*n_1*(2*n_0*n_1 - n))/(n^2*(n - 1))
Z <- (R - R_est)/sqrt(R_var)
p <- pnorm(Z)
cbind(R, R_est, R_var, Z, p)
```

```{r runs_test_graph}
xs <- data.frame(x = seq(-4, 4, length.out = 100)) %>% dplyr::mutate(density = dnorm(x))

ggplot(xs, aes(x, density)) + 
  geom_area(fill = "firebrick") + 
  gghighlight(x < Z)
```

With p = `r round(p, digits = 4)` we reject $H_0$ at the 5% level. The values are not randomly ordered.

d)

```{r runs_test_permutation}


```