---
title: "Assignment 3"
author: "Rory Sarten 301005654"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    pandoc_args: "--highlight-style=breezeDark"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(ggplot2)
library(gghighlight)
```

## Question 1

a)

A Runs Test tests a set of binary variables $X_1, ..., X_n$ to verify if the variables occur randomly.

$H_0$: variables occur randomly, i.e. knowing $X_1, ..., X_n$ does not help predict $X_{n+1}$.

$H_A$: variables are not random, i.e. knowing some part of the sequence can help predict subsequent variables.

As the variables are binary, they will take the value $0$ or $1$. The number of $0$s is $n_0$ and the number of $1$s is $n_1$, where:

$n_0 = n -\sum\limits_{i=1}^nX_i$

$n_1 = \sum\limits_{i=1}^nX_i$

To perform a Runs Test the observations are combined into one collection of $n = n_0 + n_1$ observations and arranged in increasing order of magnitude or observation. They are labeled according to which set they originally came from. A run is a group of two or more sequential values of $0$ or $1$.

Let $R$ denote the number of runs in the combined ordered sample of $X \in \{0, 1\}$. Under $H_0$, $R$ can be approximated as a normally distributed random variable, assuming both $n_0$ and $n_1$ are sufficiently large.

$R = 1 + \sum\limits_{i=2}^nI_{(X_i, X_{i-1})}$, where $I_{(X_i, X_{i-1})} = 0$ if $X_i = X_{i-1}$ and $I_{(X_i, X_{i-1})} = 1$ if $X_i \ne X_{i-1}$

$\bar{R} = \dfrac{2n_0n_1}{n} + 1$

$Var(\bar{R}) = \dfrac{2n_0n_1(2n_0n_1-n)}{n^2(n - 1)}$

With test statistic $Z = \dfrac{R-\bar{R}}{\sqrt{Var(\bar{R})}}$ where $Z \sim N(0, 1)$

b)

A small number of runs (a small value for $R$) would indicate that $X_i$ is more likely to be the same as $X_{i-1}$. A large number means that $X$ is fluctuating regularly between values and $X$ is less likely to be the same as $X_{i - 1}$.

c)

```{r runs_test_parametric}
## 0 healthy, 1 diseased
X <- "HHHDDDHDDHHHHHDHHHHHHHHHHHDDDHHHHDDDHHHHDHHHDHH" %>% 
  stringr::str_split("") %>% unlist() %>% `==`("D") %>% as.integer()

n <- length(X)
n_0 <- n - sum(X)
n_1 <- sum(X)

R <- 1 + sum(X[2:n] != X[1:n-1])
R_est <- 1 + 2*n_0*n_1/n
R_var <- (2*n_0*n_1*(2*n_0*n_1 - n))/(n^2*(n - 1))
Z <- (R - R_est)/sqrt(R_var)
p <- pnorm(Z)
cbind(R, R_est, R_var, Z, p)
```

With p-value of `r round(p, digits = 4)` we reject $H_0$ at the 5% level. We conclude that values are not randomly ordered.

d)

```{r runs_test_permutation}
set.seed(101)

calc_R <- function(input)
  1 + sum(input[2:length(input)] != input[1:length(input)-1])

sample_R <- function(iter, input)
  input %>% sample() %>% calc_R()

obs_R <- calc_R(X)

N <- 1e4
perm_R <- 1:N %>% sapply(sample_R, input = X)

p <- sum(perm_R < obs_R)/length(perm_R)
p

ggplot() + 
  geom_histogram(data = data.frame(x = perm_R[perm_R >= obs_R]), 
                 aes(x = x), binwidth = 1, fill = "white", colour = "grey") +
  geom_histogram(data = data.frame(x = perm_R[perm_R < obs_R]), 
                 aes(x = x), binwidth = 1, fill = "firebrick") +
  geom_vline(xintercept = obs_R, colour = "firebrick") +
  ggtitle("Number of Runs for permutations") +
  xlab("Runs") +
  geom_text(aes(x = 12, y = 250), label = "Runs < observed R", colour = "firebrick")
```

Because the number of runs is not a continuous variable, there is some ambiguity around whether the p-value should be calculated comparing values < or <= or even using some half point. I have decided to use < as it gives the test the highest power.

Given this we find a p-value of `r round(p, digits = 4)`. We reject $H_0$ at the 5% level. We conclude that values are not randomly ordered.

## Question 2

a)

We could use a Poisson distribution

b)

```{r Q2_b}
dataset <- data.frame(
  size = c(2, 3, 2, 7, 5, 5, 3, 2, 6, 10, 3, 2, 2, 5, 6, 4, 4, 5, 4, 4, 6, 5, 4, 2),
  village = rep(c("A", "B", "C"), c(9, 8, 7))
)

ggplot(dataset, aes(x = village, y = size)) + 
  geom_boxplot() +
  ggtitle("Family size by village") +
  ylab("family size")

## village means
tapply(dataset$size, dataset$village, mean) %>% round(digits = 4)
```

c)

```{r Q2_c}
set.seed(101)

calc_F <- function(input)
  lm(size ~ village, data = input) %>% anova() %>% .[["F value"]] %>% .[1]

sample_F <- function(iter, input)
  input %>% dplyr::mutate(village = sample(village)) %>% calc_F()

obs_F <- calc_F(dataset) %>% round(digits = 4)

N <- 1e4
perm_F <- 1:N %>% sapply(sample_F, input = dataset)

p <- sum(perm_F > obs_F)/length(perm_F) %>% round(digits = 4)
p

ggplot() + 
  geom_histogram(data = data.frame(x = perm_F), 
                 aes(x = x), bins = 50, fill = "white", colour = "grey") +
  geom_vline(xintercept = obs_F, colour = "firebrick", linetype = "dashed") +
  ggtitle("Histogram of the F Statistic") +
  xlab("F statistic") +
  geom_text(aes(x = 1, y = 150), label = "observed F", colour = "firebrick")
```

Given a p-value of `r p` we fail to reject $H_0$ and find that there is not difference in the village mean family sizes.

d)

```{r Q2_d}

test_p <- function(i = 1, input, perm = FALSE) {
  if (perm)
    input <- input %>% dplyr::mutate(village = sample(village))
  lm(size ~ village, data = input) %>% anova %>% .[["Pr(>F)"]] %>% .[1]
}

test_r2 <- function(i = 1, input, perm = FALSE) {
  if (perm)
    input <- input %>% dplyr::mutate(village = sample(village))
  lm(size ~ village, data = input) %>% summary %>% .[["r.squared"]]
}

draw_hist <- function(perm_values, observed) {
  h <- ggplot() + 
  geom_histogram(data = data.frame(x = perm_values), 
                 aes(x = x), bins = 10, fill = "white", colour = "grey") +
  geom_vline(xintercept = observed, colour = "firebrick", linetype = "dashed") +
  ggtitle("Histogram of permutation values") +
  xlab("test statistic")
  print(h)
}

permutation_test = function(test_func, dataset, niter, comparator) {
  test_stat <- test_func(input = dataset)
  perms <- 1:niter %>% sapply(test_func, input = dataset, perm = TRUE)
  
  draw_hist(perms, test_stat)
  
  sum(comparator(test_stat, perms))/niter
}

set.seed(202)
p <- permutation_test(test_p, dataset, 1e4, `>`)

set.seed(303)
p_r2 <- permutation_test(test_r2, dataset, 1e4, `<`)

```